Mirror, mirror on the wall. Which's the best match of them all?

### 2024-01-19 PNAS/index.Rnw
Tweaking conclusions, removing can be phrasing

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts using the putative tool, which can be compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire; depending on the tool design, there can be between two and four cutting surfaces in contact with the substrate.

@@

The first option gives us the maximum number of comparisons ($\ell/r - d/r + 1$), while the second option gives us the minimum number of comparisons.
In the first case, sequential comparisons share much of the same physical data and are highly related; in the second case, no data is shared between physical comparisons and we can expect that they are statistically independent, though empirically there will be nonzero correlations due to physical similarities between striations.
For simplicity, let us consider the number of comparisons to lie somewhere between these two estimates.

@@

Let us consider a wire-cutting tool with a $\Sexpr{ell/10}$ cm razor blade that meets a cast surface (one such tool is shown in \Cref{fig:tool-pic}); the wire is held against this rectangular cast surface as the blade is pushed into the wire, splitting it in two.
This is a minimal scenario - the wire will acquire striations from one side of the blade, while the blade itself has two cutting edges, which we will call side A and side B.
A blade cut of a sheet of aluminum will thus produce two striated edges corresponding to side A and side B which can be compared to cut wires to assess similarity.
We also have a 12 gauge aluminum wire ($\Sexpr{d}$ mm diameter) which may have been cut by the wire-cutting tool described above.
Class characteristics, which are shared by all tools of similar manufacture, appear to match:
there is a flat impression on one side of the wire corresponding to the cast metal backstop of the tool,
and the wire is cut such that the blade and the backstop appear to be perpendicular
(that is, the wire appears to have been cut with a tool of similar configuration).

In this example, $\ell = \Sexpr{ell}$ mm, $d = \Sexpr{d}$ mm, and at minimum there are $\ell/d = \Sexpr{ncomp_min}$ comparisons between a wire cut and a blade cut.
As there are two blade cuts (side A and side B), the minimal number of comparisons overall is \Sexpr{ncomp_min*2}, as these comparisons are non-overlapping and independent (on average).

Let's assume a resolution of $\Sexpr{res*1000} \mu$m per pixel, the maximum number of comparisons per blade cut is approximately $\Sexpr{format(round(ncomp_max/1000)*1000, big.mark=',')}$ comparisons; thus, we need $\Sexpr{format(round(ncomp_max/1000)*2000, big.mark=',')}$ comparisons overall in order to find the optimal alignment between the wire and the blade cut.
These comparisons are implicit in the calculation of cross-correlation, which is the first and often the only step used to quantitatively assess the similarity between striated evidence such as bullets, aperture shear, and firing pin impressions.
Implicit comparisons are not unique to algorithms; an examiner would need to physically align the wire and the blade cut by searching along the length of the cut to visually match striations, performing the same process physically that the algorithm performs computationally.
While these sequential comparisons are highly auto correlated, and we cannot assume sequential independence when calculating the probability of an error, they serve as an upper bound on the number of comparisons which could be performed.

@@

If we begin by considering the minimal $\ell/d$ independent comparisons between a blade cut and a wire, we can compute the overall false positive rate for all comparisons using $e$, the false positive rate for one comparison, and simple probability calculations.
Let $$E_n:=P\left(\{\geq 1\text{ false pos. in }n\text{ disjoint comparisons}\}\right).$$
Then, $E_1 = e$, $\overline{E_1} = 1-e$, $\overline{E_n} = \left[\overline{E_1}\right]^n$, and $E_n = 1 - \overline{E_n} = 1 - [1-e]^n$.
% Using this relationship between $n$ and the false positive rate $e$, we can determine what $e$ must be in order to limit $E_n$ to an acceptable level by solving for $e$: $e = 1 - \left[1 - E_n\right]^{1/n}$.
\Cref{fig:Accuracy} shows $E_n$ on the y-axis; if 90\% is an acceptable examination-wide false positive rate (FPR), 2 comparisons can be made when the single-comparison FPR is 0.05; this increases to 10 comparisons at 0.01. Thus, at an error rate of \Sexpr{sprintf("%.2f", fpr)}, examiners can conduct approximately 5 comparisons without exceeding the examination-wide FPR of 10\%.

It is clear that under these constraints, the accuracy of an examination involving multiple comparisons between a wire and a tool is likely to be low, as the number of candidate alignments that must be examined is too high.
Even the most innocuous example we could find (small blade, only 2 cutting surfaces, and a relatively large wire) involved a minimum of \Sexpr{2*ncomp_min} completely independent comparisons.
As a result, findings from  wire comparisons made under current protocols (using error rates recently published in the forensics literature) are probably not reliable enough to be presented at trial.
When we then consider that examiners would make cuts at multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons even in the simple case we have presented here, the probability of a false positive error increases further still.

@@

Forensic practitioners often report the findings from their examinations in the form of a categorical conclusion reflecting a single decision. This can be misleading when this decision relies on multiple comparisons which are not individually presented in reports or testimony. In this short contribution, we have shown that the implicit comparisons performed during the forensic analysis of wire cuts have the potential to vastly increase the overall chance that at least one false positive error occurs.
%While often a forensic conclusion is presented as a single decision, underlying that decision are many more comparisons which are not discussed or reported upon.
Omission of the underlying details is not necessarily due to malice; rather, it results from protocols which focus on finding the best possible match between two items and then evaluating that match, and also from lack of familiarity with basic statistical ideas.

In our example we highlight a simple case where a wire  is cut using a two-sided blade, but the findings apply to many other situations where a forensic evaluation involves multiple comparisons against a set of accumulated information, such as database searches.
%either in a database or in a series of test samples created during the examination process.
It is critically important that forensic practitioners become aware of the impact of carrying out multiple comparisons on their final conclusions and that lab procedures and testimony are modified accordingly.
We suggest three strategies to improve transparency of forensic examinations and to produce more reliable estimates of examination-wide error rates.

>>>>

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts using the putative tool, which are then compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire; depending on the tool design, there can be between two and four cutting surfaces in contact with the substrate.

@@

The first option gives us the maximum number of comparisons ($\ell/r - d/r + 1$), while the second option gives us the minimum number of comparisons $\ell/d$.
In the first case, sequential comparisons share much of the same physical data and are highly related; in the second case, no data is shared between physical comparisons and we can expect that they are statistically independent, though empirically there will be nonzero correlations due to physical similarities between striations.
For simplicity, let us consider the number of comparisons to lie somewhere between these two estimates.
Note that when $\ell/d \approx 1$, as in bullet LEA comparisons, the minimal and the maximal number of comparisons are both close to 1.

@@

Let us consider a wire-cutting tool with a $\Sexpr{ell/10}$ cm razor blade that meets a cast surface (one such tool is shown in \Cref{fig:tool-pic}); the wire is held against this rectangular cast surface as the blade is pushed into the wire, splitting it in two.
This is a minimal scenario - the wire will acquire striations from one side of the blade, while the blade itself has two cutting edges, which we will call side A and side B.
A blade cut of a sheet of aluminum will thus produce two striated edges corresponding to side A and side B which are compared to cut wires to assess similarity.
We also have a 12 gauge aluminum wire ($\Sexpr{d}$ mm diameter) which may have been cut by the wire-cutting tool described above.
Class characteristics, which are shared by all tools of similar manufacture, appear to match:
there is a flat impression on one side of the wire corresponding to the cast metal backstop of the tool,
and the wire is cut such that the blade and the backstop appear to be perpendicular
(that is, the wire appears to have been cut with a tool of similar configuration).

In this example, $\ell = \Sexpr{ell}$ mm, $d = \Sexpr{d}$ mm, and there are at least $\ell/d = \Sexpr{ncomp_min}$ comparisons between a wire cut and a blade cut.
As there are two blade cuts (side A and side B), the minimal number of comparisons overall is \Sexpr{ncomp_min*2}, as these comparisons are non-overlapping and independent (on average).

Assuming a resolution of $\Sexpr{res*1000} \mu$m per pixel, the maximum number of comparisons per blade cut is approximately $\Sexpr{format(round(ncomp_max/1000)*1000, big.mark=',')}$ comparisons; thus, we need $\Sexpr{format(round(ncomp_max/1000)*2000, big.mark=',')}$ comparisons overall in order to find the optimal alignment between the wire and the blade cut.
These comparisons are implicit in the calculation of cross-correlation, which is the first and often the only step used to quantitatively assess the similarity between striated evidence such as bullets, aperture shear, and firing pin impressions.
Implicit comparisons are not unique to algorithms; an examiner would need to physically align the wire and the blade cut by searching along the length of the cut to visually match striations, performing the same process physically that the algorithm performs computationally.
While these sequential comparisons are highly auto correlated, and we cannot assume sequential independence when calculating the probability of an error, they serve as an upper bound on the number of comparisons which could be performed.

As the number of comparisons increases, the probability of encountering a coincidental match increases.
Statisticians call this the \textit{familywise error rate} ("family" = number of comparisons); it is an important quantity to control when conducting a series of tests.

@@

For a single-comparison FPR of $e$, the familywise FPR for $n$ comparisons is $1 - [1-e]^n$.
\Cref{fig:Accuracy} shows the familywise accuracy rate on the y-axis; if 90\% is an acceptable overall accuracy, 2 comparisons can be made when $e = 0.05$; this increases to 10 comparisons when $e =  0.01$.
Thus, at an error rate of \Sexpr{sprintf("e = %.2f", fpr)}, examiners can conduct approximately 5 comparisons without exceeding the familywise FPR of 10\%.

It is clear that under these constraints, the accuracy of an examination involving multiple comparisons between a wire and a tool will be low, as the number of candidate alignments that must be examined is high.
Even the most innocuous example (small blade, only 2 cutting surfaces, and a relatively large wire) involved a minimum of \Sexpr{2*ncomp_min} completely independent comparisons.
When we then consider that examiners would make cuts at multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons even in the simple case we have presented here, the probability of a false identification becomes even more likely.
As a result, it is questionable whether wire comparisons made under current protocols are reliable enough to be presented at trial.

@@

Forensic practitioners often report the findings from their examinations in the form of a categorical conclusion reflecting a single decision.
This is misleading when the decision relies on multiple comparisons which are not individually presented in reports or testimony.
In this short contribution, we have shown that the implicit comparisons performed during forensic analysis of wire cuts increase the familywise error rate.
Omission of the underlying details is not necessarily due to malice; rather, it results from protocols which focus on finding the best possible match between two items and evaluating that match.

In our example we highlight a simple case where a wire is cut using a two-sided blade, but the findings apply to many other situations where a forensic evaluation involves multiple comparisons against a set of accumulated information, such as database searches.
It is critically important that forensic practitioners become aware of the impact of carrying out multiple comparisons on their final conclusions and that lab procedures and testimony are modified accordingly.
We suggest three strategies to improve transparency of forensic examinations and to produce more reliable estimates of examination-wide error rates.
